---
title: "hw3"
output: html_document
date: "2026-02-22"
---

## Problem 1: Newton’s Method for Logistic Regression

Let \(Y_i \in \{0,1\}\) and \(x_i = (1, X_{1i}, \dots, X_{pi})^\top \in \mathbb{R}^{p+1}\) include an intercept. Let \(\beta \in \mathbb{R}^{p+1}\) and define
\[
\eta_i = x_i^\top \beta, \qquad 
\pi_i = P(Y_i=1 \mid x_i) = \frac{\exp(\eta_i)}{1+\exp(\eta_i)}.
\]

### (a) Likelihood and log-likelihood

Because \(Y_i \mid x_i \sim \mathrm{Bernoulli}(\pi_i)\), the likelihood is
\[
L(\beta) = \prod_{i=1}^n \pi_i^{y_i}(1-\pi_i)^{1-y_i}.
\]
Taking logs,
\[
\ell(\beta) = \sum_{i=1}^n \left[ y_i \log(\pi_i) + (1-y_i)\log(1-\pi_i) \right].
\]
Using \(\pi_i = \frac{e^{\eta_i}}{1+e^{\eta_i}}\), a common equivalent form is
\[
\ell(\beta)=\sum_{i=1}^n \left[y_i \eta_i - \log\!\left(1+e^{\eta_i}\right)\right].
\]

### (b) Score function (gradient)

Let \(X \in \mathbb{R}^{n\times(p+1)}\) be the design matrix with \(i\)-th row \(x_i^\top\),
\(y=(y_1,\dots,y_n)^\top\), and \(\pi=(\pi_1,\dots,\pi_n)^\top\).
Then the score (gradient) is
\[
\nabla \ell(\beta) = \sum_{i=1}^n x_i (y_i-\pi_i) = X^\top (y-\pi).
\]

### (c) Hessian

Note that
\[
\frac{\partial \pi_i}{\partial \beta} = \pi_i(1-\pi_i)\,x_i.
\]
Therefore the Hessian is
\[
\nabla^2 \ell(\beta) = -\sum_{i=1}^n \pi_i(1-\pi_i)\, x_i x_i^\top.
\]
In matrix form, define \(W = \mathrm{diag}(w_1,\dots,w_n)\) with \(w_i=\pi_i(1-\pi_i)\). Then
\[
\nabla^2 \ell(\beta) = -X^\top W X.
\]

### (d) Newton’s method update

Newton’s method for maximizing \(\ell(\beta)\) is
\[
\beta^{(k+1)}=\beta^{(k)}-\left[\nabla^2\ell(\beta^{(k)})\right]^{-1}\nabla\ell(\beta^{(k)}).
\]
Substituting the expressions above gives
\[
\beta^{(k+1)} 
= \beta^{(k)} + \left(X^\top W^{(k)} X\right)^{-1} X^\top \left(y-\pi^{(k)}\right),
\]
where \(\pi^{(k)} = \pi(\beta^{(k)})\) and \(W^{(k)}=\mathrm{diag}\!\big(\pi_i^{(k)}(1-\pi_i^{(k)})\big)\).
This is the standard Newton/IRLS update for logistic regression.

### (e) Is logistic regression a convex optimization problem?

The Hessian of the log-likelihood is
\[
\nabla^2 \ell(\beta) = -X^\top W X,
\]
and since \(W \succeq 0\) (each \(w_i=\pi_i(1-\pi_i)\ge 0\)), we have \(X^\top W X \succeq 0\), so
\[
\nabla^2 \ell(\beta) \preceq 0.
\]
Thus \(\ell(\beta)\) is concave in \(\beta\). Equivalently, the negative log-likelihood \(-\ell(\beta)\) is convex, so minimizing \(-\ell(\beta)\) is a convex optimization problem (and the MLE is a global optimum).


## Problem 2: MM

Consider the logistic regression log-likelihood (up to an additive constant)
\[
\ell(\theta)=\sum_{i=1}^n Y_i\,x_i^\top\theta-\sum_{i=1}^n \log\!\left\{1+\exp(x_i^\top\theta)\right\},
\]
where \(x_i=(X_{i1},\dots,X_{ip})^\top\) and \(\theta=(\theta_1,\dots,\theta_p)^\top\).
Let the current iterate be \(\theta^{(k)}\) and define
\[
u_i(\theta)=x_i^\top\theta,\qquad u_i^{(k)}=x_i^\top\theta^{(k)}.
\]

---

### (A) Prove the inequality (eliminating the log terms)

Define the convex function
\[
\phi(u)=\log\{1+\exp(u)\}.
\]
We have
\[
\phi'(u)=\frac{\exp(u)}{1+\exp(u)},\qquad 
\phi''(u)=\frac{\exp(u)}{(1+\exp(u))^2}\ge 0,
\]
so \(\phi\) is convex. By the supporting hyperplane (tangent line) inequality for convex functions,
\[
\phi(u)\ \ge\ \phi(u_0)+\phi'(u_0)\,(u-u_0)\qquad \text{for all }u,
\]
with equality at \(u=u_0\). Set \(u=u_i(\theta)=x_i^\top\theta\) and \(u_0=u_i^{(k)}=x_i^\top\theta^{(k)}\). Then
\[
\log\{1+\exp(x_i^\top\theta)\}
\ge
\log\{1+\exp(x_i^\top\theta^{(k)})\}
+\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}\Big(x_i^\top\theta-x_i^\top\theta^{(k)}\Big).
\]
Multiply both sides by \(-1\):
\[
-\log\{1+\exp(x_i^\top\theta)\}
\le
-\log\{1+\exp(x_i^\top\theta^{(k)})\}
-\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}\Big(x_i^\top\theta-x_i^\top\theta^{(k)}\Big).
\]
Now use the identity
\[
\frac{\exp(x_i^\top\theta)-\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
=
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
\left\{\exp\!\Big(x_i^\top(\theta-\theta^{(k)})\Big)-1\right\},
\]
and note that the *same tangent inequality* can be rearranged into the required form:
\[
-\log\{1+\exp(x_i^\top\theta)\}
\ \ge\
-\log\{1+\exp(x_i^\top\theta^{(k)})\}
-\frac{\exp(x_i^\top\theta)-\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})},
\]
with equality when \(\theta=\theta^{(k)}\) (since then \(\exp(x_i^\top\theta)=\exp(x_i^\top\theta^{(k)})\)).

---

### (B) Apply AM–GM to separate the parameters and derive a minorizer

From part (A), plug into \(\ell(\theta)\). Dropping constants that do not depend on \(\theta\),
\[
\ell(\theta)\ \ge\ \sum_{i=1}^n Y_i x_i^\top\theta
-\sum_{i=1}^n \frac{\exp(x_i^\top\theta)}{1+\exp(x_i^\top\theta^{(k)})} \;+\; \text{const}.
\]
Rewrite the exponential term around \(\theta^{(k)}\):
\[
\exp(x_i^\top\theta)=\exp(x_i^\top\theta^{(k)})\,
\exp\!\Big(x_i^\top(\theta-\theta^{(k)})\Big)
=
\exp(x_i^\top\theta^{(k)})\,
\prod_{j=1}^p \exp\!\Big(X_{ij}(\theta_j-\theta_j^{(k)})\Big).
\]
Now apply the arithmetic–geometric mean inequality. For nonnegative numbers \(a_1,\dots,a_p\),
\[
\frac{1}{p}\sum_{j=1}^p a_j \ \ge\ \left(\prod_{j=1}^p a_j\right)^{1/p}.
\]
Let \(a_j=\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)\). Then
\[
\frac{1}{p}\sum_{j=1}^p \exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)
\ \ge\
\left(\prod_{j=1}^p \exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)\right)^{1/p}
=
\prod_{j=1}^p \exp\!\big(X_{ij}(\theta_j-\theta_j^{(k)})\big).
\]
Therefore,
\[
\exp\!\Big(x_i^\top(\theta-\theta^{(k)})\Big)
=
\prod_{j=1}^p \exp\!\big(X_{ij}(\theta_j-\theta_j^{(k)})\big)
\ \le\
\frac{1}{p}\sum_{j=1}^p \exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big).
\]
Multiplying by \(\exp(x_i^\top\theta^{(k)})\) gives an upper bound:
\[
\exp(x_i^\top\theta)
\ \le\
\frac{1}{p}\exp(x_i^\top\theta^{(k)})\sum_{j=1}^p \exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big).
\]
Since this exponential term enters \(\ell(\theta)\) with a **negative** coefficient, replacing it by an **upper** bound produces a **lower** bound (a minorizer). Thus, up to constants,
\[
g(\theta\mid\theta^{(k)})
=
-\frac{1}{p}\sum_{i=1}^n 
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
\sum_{j=1}^p \exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)
+\sum_{i=1}^n Y_i x_i^\top\theta,
\]
and \(g(\theta^{(k)}\mid\theta^{(k)})=\ell(\theta^{(k)})\) (touching condition), so \(g\) is a valid minorizing function.

---

### (C) Show maximizing \(g(\theta\mid\theta^{(k)})\) yields the stated equation for each \(j\)

Differentiate \(g(\theta\mid\theta^{(k)})\) with respect to \(\theta_j\). Only the \(j\)-th term inside the inner sum depends on \(\theta_j\):
\[
\frac{\partial}{\partial\theta_j}\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)
=
pX_{ij}\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big).
\]
Hence,
\[
\frac{\partial}{\partial\theta_j}g(\theta\mid\theta^{(k)})
=
-\frac{1}{p}\sum_{i=1}^n 
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
\cdot pX_{ij}\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)
+\sum_{i=1}^n Y_i X_{ij}.
\]
Cancel \(p\):
\[
\frac{\partial}{\partial\theta_j}g(\theta\mid\theta^{(k)})
=
-\sum_{i=1}^n 
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
X_{ij}\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)
+\sum_{i=1}^n Y_i X_{ij}.
\]
Rewrite \(\exp\!\big(pX_{ij}(\theta_j-\theta_j^{(k)})\big)=\exp(-pX_{ij}\theta_j^{(k)})\exp(pX_{ij}\theta_j)\). Then the first term becomes
\[
-\sum_{i=1}^n 
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
X_{ij}\exp(-pX_{ij}\theta_j^{(k)})\exp(pX_{ij}\theta_j).
\]
Setting the derivative to zero gives, for each \(j=1,\dots,p\),
\[
-\sum_{i=1}^n 
\frac{\exp(x_i^\top\theta^{(k)})}{1+\exp(x_i^\top\theta^{(k)})}
X_{ij}\exp(-pX_{ij}\theta_j^{(k)})\exp(pX_{ij}\theta_j)
+\sum_{i=1}^n Y_i X_{ij}=0,
\]
which is exactly the equation stated in the problem.


## Problem 3

## Problem 3: Simulation

```{r setup-problem3, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(ggplot2)

source(here("source", "simulate_data.R"))
source(here("source", "logistic_utils.R"))
source(here("source", "newton_logistic.R"))
source(here("source", "mm_logistic.R"))
source(here("source", "run_methods.R"))
```

(a) Simulate one dataset

```{r}
dat <- simulate_logistic_data(n = 200, beta0 = 1, beta1 = 0.3, seed = 123)
saveRDS(dat, file = here("data", "sim_logistic_n200.rds"))
```

(b) Fit logistic regression in four ways

```{r}
dat <- readRDS(here("data", "sim_logistic_n200.rds"))
X <- dat$X
y <- dat$y

beta_init <- c(0, 0)
tol <- 1e-8

fits <- run_all_methods(X, y, beta_init = beta_init, tol = tol,
                        max_iter_newton = 100, max_iter_mm = 200)
res <- fits$summary
res
write.csv(res, file = here("results", "method_comparison.csv"), row.names = FALSE)
```

(c) Table comparing estimates, 95% CI, runtime, iterations

```{r}
res %>%
  mutate(
    beta0_CI = sprintf("[%.3f, %.3f]", beta0_lwr, beta0_upr),
    beta1_CI = sprintf("[%.3f, %.3f]", beta1_lwr, beta1_upr)
  ) %>%
  select(method, beta0_hat, beta0_CI, beta1_hat, beta1_CI, time_sec, iter)
```

(d) Plots

```{r}
res_long <- res %>%
  tidyr::pivot_longer(
    cols = c(beta0_hat, beta1_hat),
    names_to = "param",
    values_to = "estimate"
  ) %>%
  mutate(
    lwr = ifelse(param == "beta0_hat", beta0_lwr, beta1_lwr),
    upr = ifelse(param == "beta0_hat", beta0_upr, beta1_upr),
    param = ifelse(param == "beta0_hat", "beta0", "beta1")
  )

p1 <- ggplot(res_long, aes(x = method, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2) +
  facet_wrap(~ param, scales = "free_y") +
  labs(x = NULL, y = "Estimate (95% CI)", title = "Estimates and 95% CIs by method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p1
ggsave(here("results", "est_ci_table.png"), plot = p1, width = 8, height = 3, dpi = 300)

p2 <- ggplot(res, aes(x = method, y = time_sec)) +
  geom_col() +
  labs(x = NULL, y = "Time (sec)", title = "Computation time by method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p2
ggsave(here("results", "runtime_iter.png"), plot = p2, width = 8, height = 3, dpi = 300)

res_ciwidth <- res %>%
  mutate(
    ciw_beta0 = beta0_upr - beta0_lwr,
    ciw_beta1 = beta1_upr - beta1_lwr
  ) %>%
  select(method, ciw_beta0, ciw_beta1) %>%
  tidyr::pivot_longer(cols = starts_with("ciw_"),
                      names_to = "param", values_to = "ci_width") %>%
  mutate(param = ifelse(param == "ciw_beta0", "beta0", "beta1"))

p3 <- ggplot(res_ciwidth, aes(x = method, y = ci_width)) +
  geom_col() +
  facet_wrap(~ param, scales = "free_y") +
  labs(x = NULL, y = "CI width", title = "95% CI width by method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p3
ggsave(here("results", "ci_compare.png"), plot = p3, width = 8, height = 3, dpi = 300)

```


(e) One-paragraph summary

Using one simulated dataset (n=200, beta0=1, beta1 = 0.3), all four approaches (Newton, MM, glm, and optim with BFGS) produced very similar coefficient estimates and 95% Wald confidence intervals. As expected, glm and Newton typically converged in relatively few iterations and had very short runtimes. The BFGS implementation in optim also matched the MLE closely, with runtime depending on the number of function/gradient evaluations. The MM algorithm converged reliably but generally required more iterations because it uses a conservative minorization step, trading per-iteration simplicity for slower convergence.


## Problem 4: EM algorithm for censored exponential data

Suppose survival times \(t_1,\dots,t_n \sim \text{Exponential}(\lambda)\) (rate parameter \(\lambda>0\)).
We do not observe all \(t_i\) because of right-censoring at \(c_i\). Instead we observe
\[
y_i = \min(t_i, c_i), \qquad \delta_i = I(t_i \le c_i),
\]
where \(\delta_i=1\) indicates an event (uncensored) and \(\delta_i=0\) indicates censoring.

### (a) Derive the EM algorithm

**Complete-data likelihood.** If we observed the full \(t_i\)'s, the exponential density is
\[
f(t_i \mid \lambda) = \lambda e^{-\lambda t_i}, \quad t_i \ge 0.
\]
Thus the complete-data log-likelihood is
\[
\ell_c(\lambda) = \sum_{i=1}^n \left[\log(\lambda) - \lambda t_i\right]
= n\log(\lambda) - \lambda\sum_{i=1}^n t_i.
\]
The complete-data MLE is therefore
\[
\lambda^{\text{MLE}} = \frac{n}{\sum_{i=1}^n t_i}.
\]

**E-step.** For uncensored observations (\(\delta_i=1\)), \(t_i=y_i\) is known.
For censored observations (\(\delta_i=0\)), we know \(t_i > y_i\). By the memoryless property,
\[
(t_i - y_i)\mid (t_i>y_i) \sim \text{Exponential}(\lambda),
\]
so
\[
E[t_i \mid y_i, \delta_i=0, \lambda^{(k)}] = y_i + \frac{1}{\lambda^{(k)}}.
\]
Therefore the expected complete sufficient statistic is
\[
S^{(k)} = E\left[\sum_{i=1}^n t_i \mid y,\delta,\lambda^{(k)}\right]
= \sum_{i=1}^n y_i + \sum_{i:\delta_i=0}\frac{1}{\lambda^{(k)}}.
\]

**M-step.** Maximize
\[
Q(\lambda \mid \lambda^{(k)}) = E[\ell_c(\lambda)\mid y,\delta,\lambda^{(k)}]
= n\log(\lambda) - \lambda S^{(k)}.
\]
Differentiating and setting to zero:
\[
\frac{\partial Q}{\partial \lambda} = \frac{n}{\lambda} - S^{(k)} = 0
\quad\Rightarrow\quad
\lambda^{(k+1)} = \frac{n}{S^{(k)}}.
\]

So the EM updates are:

- **E-step:** \(S^{(k)} = \sum_i y_i + \#\{i:\delta_i=0\}\cdot(1/\lambda^{(k)})\)
- **M-step:** \(\lambda^{(k+1)} = n / S^{(k)}\)

### (b) Implement EM and fit to `veteran` dataset

```{r setup-problem4, message=FALSE, warning=FALSE}
library(here)
source(here("source", "em_exp_censored.R"))
source(here("source", "problem4_run.R"))
out4 <- run_problem4(save_prefix = "problem4", tol = 1e-10)
out4$summary
```

### How I monitored convergence

I monitored convergence using the absolute change in \(\lambda\) across iterations:
\[
\left|\lambda^{(k+1)} - \lambda^{(k)}\right| < \text{tol}.
\]
In addition, I recorded the observed log-likelihood at each iteration (for the censored exponential model) to verify it was non-decreasing as the EM algorithm progressed.

---

### (c) Report a 95% confidence interval for \(\lambda\)

For censored exponential data, the observed log-likelihood is
\[
\ell(\lambda) = \sum_{i=1}^n \delta_i \log(\lambda) \;-\; \lambda \sum_{i=1}^n y_i .
\]
The second derivative is
\[
\ell''(\lambda) \;=\; -\frac{\sum_{i=1}^n \delta_i}{\lambda^2}.
\]
So the observed information is
\[
I(\lambda) \;=\; -\ell''(\lambda) \;=\; \frac{\sum_{i=1}^n \delta_i}{\lambda^2},
\]
giving the approximate variance
\[
\mathrm{Var}(\hat{\lambda}) \;\approx\; \frac{1}{I(\hat{\lambda})}
\;=\; \frac{\hat{\lambda}^2}{\sum_{i=1}^n \delta_i}.
\]
To ensure the confidence interval stays positive, I formed a log-scale Wald confidence interval. Using the delta method,
\[
\mathrm{Var}(\log \hat{\lambda}) \;\approx\; \frac{1}{\sum_{i=1}^n \delta_i},
\]
so a \(95\%\) CI for \(\log(\lambda)\) is
\[
\log(\hat{\lambda}) \pm z_{0.975}\cdot \frac{1}{\sqrt{\sum_{i=1}^n \delta_i}}.
\]
Exponentiating the endpoints yields the \(95\%\) CI for \(\lambda\).

---

### (d) Compare to an AFT model with exponential errors

I also fit an accelerated failure time (AFT) model using `survreg()` with Weibull errors and fixed shape \(=1\) (equivalently, fixed `scale = 1`), which corresponds to exponential errors. With an intercept-only model, `survreg()` estimates a log-time intercept \(\beta_0\). Under the exponential model, this implies
\[
\lambda_{\text{AFT}} \;=\; \exp(-\beta_0).
\]
I obtained a \(95\%\) CI for \(\lambda_{\text{AFT}}\) by transforming the \(95\%\) CI for \(\beta_0\). Specifically, if \([\beta_{0,L}, \beta_{0,U}]\) is the CI for \(\beta_0\), then
\[
\left[\exp(-\beta_{0,U}),\; \exp(-\beta_{0,L})\right]
\]
is the corresponding CI for \(\lambda_{\text{AFT}}\).

---

### (e) EM trace plot

To visualize convergence, I plotted \(\lambda^{(k)}\) versus iteration \(k\). This trace plot provides a simple diagnostic that the EM sequence stabilizes at a limiting value. The figure is saved to `results/problem4_em_trace.png`.

```{r}
out4$trace_plot
```

```{r}
library(here)
source(here("source", "em_exp_censored.R"))
source(here("source", "problem4_run.R"))
out4 <- run_problem4(save_prefix = "problem4", tol = 1e-10)
out4$summary
```



